---
title: Convert all Character variables to Factors
author: Michael Barrowman
date: '2020-10-15'
slug: convert-all-character-variables-to-factors
categories: []
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2020-10-15T15:05:08+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

# Introduction

First, let's load up what we need!

```{r LoadAndSeed, echo=T, results='hide', message=F}
set.seed(15102020)
library(tidyverse) #We'll use tidyverse functions
library(magrittr) #A few extra pipes from magrittr
library(lexicon) #For a word dictionary
```

When dealing wth complex datasets, it is common that a variable may be stored as a *character* variable, when in reality what you want is a *factor* variable. On the surface, these two constructs look very similar:

```{r EgFactorvsCharacter}
eg_df <- tibble(
  c_var = c("Cat","Dog","Cat","Mouse","Mouse"),
  f_var = factor(c("Cat","Dog","Cat","Mouse","Mouse"))
)
eg_df
```

However, underneath they are treated quite differently. Behind the scenes, the factors are actually stored as integers with a special lookup table called their _levels_, which can be seen if we print the variable individually:
```{r EgFactorBreakdown}
eg_df$f_var
```

We can also see the hidden numbers by converting this to numeric:
```{r EgFactorNumeric}
as.numeric(eg_df$f_var)
```

The first element, `Cat` is associated with the first level, so it is stored as a `1`, the third element is also `Cat`, so it is also stored as a `1`. The fourth & fifth are both `Mouse` and so they're stored as `3`, indicating to use the third level.

# Why factors?

Most statistical operations within R that can act on a *character* variable will essentially convert to a *factor* first. So, it's more efficient to convert characters to factors before passing them into these kinds of functions. This also gives us more control over what we're going to get.

This conversion makes many processes that work with characters a bit slow. If you're wanting to do 20 functions on a data set and each one needs to convert your characters to factors internally before doing what it needs to, it's clearly much faster to manually convert _once_ before using these functions.

Factors also take up slightly less space in your system's memory. In R, this is approximately half the space of a character, however the way R stores this kind of data is surprisingly efficient. It's definitely a good habit to get into if you ever want to move onto less efficient storage methods.

# Converting

Above, I used the `factor()` function to quickly convert a single character variable to a factor variable. But what about if you've got a large dataset with many, many character variables that you want to convert to factors. What's the smoothest way to do this?

## Example random dataset

First, let's create a large dataset, we'll loop through a bunch of columns. We'll use Fry's 1000 Most Commonly Use English Words, as found in the `sw_fry_1000` dataset from the `{lexicon}` package to choose random words for each variable. We'll also throw in some numeric variables to make things harder:
```{r CreateDataset}
df <- tibble(id=1:1000) #declare a tibble with just an id variable
for(i in 1:10)
{
  #How many distinct words should this variable have?
  distinct_words <- round(rexp(1,1/20)) +1
  
  #What words can we choose from for this variable?
  these_words <- sample(sw_fry_1000,distinct_words)
  
  #What's the name of this variable?
  this_name <- paste0("var_",ncol(df) + 1)
  
  #Generate the variable
  this_variable <- sample(these_words,1000,replace=T)
  
  #Store it in the tibble
  df[[this_name]] <- this_variable
  
  #Approximated 1/3 of the time, we'll add a numeric variable
  if(rbinom(1,1,1/3) == 1){
    this_name <- paste0("var_",ncol(df)+1)
    
    df[[this_name]] <- rnorm(1000)
  }
  
}
df
```

The generation of this data is actually rather clunky as it's using a loop, and we're going to avoid that. Instead, we're going to turn all these characters into factors in a single line. Here's the line of code which will update the dataset, followed by the explanation:

# The solution

With `{tidyverse}` processes, the key thing we're trying to do is build a "sentence" explaining what we're doing. Here's our expression, followed by the English sentence equivalent

```{r SimpleSolution}
df %<>% mutate(across(where(is.character),as_factor))
#Update the df by mutating it across variables where it is a
#   character by performing as_factor on them

df
```

And as if by magic, all of the characters are now factors (note the `<fct>` under the variable names).

# The Explanation

The above code uses five functions, and an operation to perform the action. We'll dig down into the functions and then climb back out as their results are processed:

* `%<>%` grabs the tibble on it's left hand side and passes it to the function on the right. At this point, it works exactly like the regular `%>%` operator
    * `mutate()` means we are creating or updating a variable inside the tibble
        * `across()` allows us perform a function _across_ many variables within the tibble
            * `where()` allows us to specify _where_ we want `across()` to perform the function
                * `is.character()`, in the above line, we don't use the brackets for `is.character()` because we're not applying it, we're referencing it. We're telling the `where()` function to use this when checking where we want the function to be applied. The `is.character()` function returned `TRUE` when the variable is a character and `FALSE` when it isn't (e.g. a numeric)
            * `where()` therefore applies this function to every variable in `df` and returns a vector of `TRUE` and `FALSE` to `across()` to indicate which variables in the tibble we want `across()` to act on
            * `as_factor()` converts things (e.g. characters) into factors.
        * `across()` has now been passed a logical vector telling it which columns to apply a function and a function that it needs to apply. So it does just that and outputs another tibble
    * `mutate()` has then been passed a tibble for it's first argument (`df` via the `%<>%` pipe) and another tibble as the output of `across()`. It stitches these together, if there are any names in common, it overwrites those in `df` with those from `across()`. All the variables in `across()` will also appear in `df` because that's where they came from, so the old values are overwritten with the new ones
* `%<>%` then receives this new tibble from `mutate()` and stores it back into the `df` tibble that we originally passed to it. This is essentially saying that `df %<>% f()` is the same as `df <- df %>% f()`, that's why this is called the _assignment pipe_ or _updating pipe_.