---
title: Getting a variable name in a pipeline
author: ~
date: '2020-07-24'
slug: getting-a-variable-name-in-a-pipeline
categories: []
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2020-07-24T00:25:35+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r, include=F}
library(rlang)
library(magrittr)
```

There is no greater staple of the `{tidyverse}` than the  pipe, `%>%`, however not a lot of people understand what's going on "under-the-bonnet" of the pipe. To be fair, not many people have to worry about it. Until you start trying to do a bit of meta-programming. Then things can get difficult.

Recently, a [question](https://community.rstudio.com/t/getting-argument-name-when-passed-via-i-keep-getting/74086) posed by user [preposterior](https://community.rstudio.com/u/preposterior) on RStudio Community embodied an issue that can happen when trying to extract the name of a variable.

When using the `{rlang}` package, we can get the name of a variable passed into a function using the following:
```{r, simple_get_name_def}
simple_get_name <- function(x){
  x_sym <- ensym(x)
  as_name(x_sym)
}

my_variable <- 1

simple_get_name(my_variable)
```

However, when we try to use this with a pipe, it goes wrong:

```{r simple_get_name_pipe}
my_variable %>% simple_get_name()
```

What is this!? Where did that dot come from? Well, that's why you're here. The `.` can actually be used as a variable name in R, it's perfectly syntactic, although ill-advised. Many functions (particularly `{tidyverse}` ones) use the `.` as a filler for other purposes, the pipe is a big example, but it is also prominent in the `map()` family of functions in `{purrr}`.

The reason we get this output is because the pipe actually turns your pipeline of functions into a chain of new functions defined something like this:
```{r anonymous_simple_get_name, eval=F}
function(.)
simple_get_name(.)
```
So, it's a chain of wrapper functions around your pipeline'd functions. The functions are direct copies of what you use in your pipeline. For example, if you have a few arguments, the `.` will be inserted as the first argument. This makes sense as this is what the pipe does (passes your input into the first agrument, unless told otherwise).

Once the pipe has this chain of functions, it uses the `freduce()` function to apply the functions in order to the output of the previous one. You already knew what a pipeline did, know you've got a little insight into how.

So how do we pull out that `my_variable` name from within a piped function? Well the problem is that, within the context of that function, that variable is lost. It's value has been put into the variable `.`, but that original name is long gone.

We can, however, look back over the call-stack where the current function is being evaluated (which is what error-finding functions like `traceback()` do). Within the pipe, it actually creates a relatively deeply nested set of calls (about 9 calls deep). However, the `sys.calls()` function can return this stack. Compare for example the following two outputs:
```{r stack_fun_dummy,eval=F}
stack_fun <- function(x){
  sys.calls()
}
stack_fun(my_variable)
my_variable %>% stack_fun
```
```{r stack_fun_def, echo=F}
stack_fun <- function(x){
  sys.calls()[-(1:24)]
}
stack_fun(my_variable)
my_variable %>% stack_fun
```

The first element of this stack will be the initial call, in this case `my_variable %>% stack_fun()`. This will be a `call` object and so we can pull out the left-hand side by extracting the second element (the `%>%` is the first element, and `stack_fun` is the third). Therefore, the previous function can be written as:
```{r stacked_get_name_dummy, eval=F}
stacked_get_name <- function(x){
  first_call <- sys.calls()[[1]] #get the first entry on the call stack
  lhs <- first_call[[2]] #get the second element of this entry
  z <- rlang::as_name(lhs)
  print(z)
}

my_variable %>% stacked_get_name()
```
```{r stacked_get_name_def, echo=F}
stacked_get_name <- function(x){
  first_call <- sys.calls()[[25]] #get the first entry on the call stack
  lhs <- first_call[[2]] #get the second element of this entry
  z <- rlang::as_name(lhs)
  print(z)
}

my_variable %>% stacked_get_name()
```
It worked! Brilliant!

But, that's not the end of our tale!

This is just looking for the initial call, and isn't strictly going to seek out where there is a pipe.  For example, it wouldn't work with the following function, since `wrap_stacked_get_name()` would be at the top of the stack:
```{r wrap_stacked_get_name_def}
wrap_stacked_get_name <- function(wrapped_var){
  this_variable <- wrapped_var+1
  this_variable %>% stacked_get_name
}

wrap_stacked_get_name(my_variable)

```
This should return `this_variable`, but since it's looking at the initial call, it looks too far back up the call-stack and misses this variable.

However, by inspecting the entire stack for a pipe, we can pull out the most recent (i.e. the lowest) entry that is a pipe, and grab the left-hand side of that call.
```{r get_lhs_def}
get_lhs <- function(){
  calls <- sys.calls()
  
  #pull out the function or operator (e.g. the `%>%`)
  call_firsts <- lapply(calls,`[[`,1) 
  
  #check which ones are equal to the pipe
  pipe_calls <- vapply(call_firsts,identical,logical(1),quote(`%>%`))
  
  #if we have no pipes, then get_lhs() was called incorrectly
  if(all(!pipe_calls)){
    NULL
  } else {
    #Get the most recent pipe, lowest on the 
    pipe_calls <- which(pipe_calls)
    pipe_calls <- pipe_calls[length(pipe_calls)]
    
    #Get the second element of the pipe call
    this_call <- calls[[c(pipe_calls,2)]]
    
    #We need to dig down into the call to find the original
    while(is.call(this_call) && identical(this_call[[1]],quote(`%>%`))){
      this_call <- this_call[[2]]
    }
    this_call
    
  }
}
```

Once we have the call, getting the lhs of it requires digging down. If we have pipeline, then it's actually a nested sequence of operators. For example, `2+3+4` makes sense to us, but R can't add like this, it breaks this down by calculating from left to right, basically it does this `(2 + 3) + 4`, which is the same as `add(add(2,3),4)`. R does this with the pipe too.

If we're piping a few things together, we write this: `my_variable %>% fun1 %>% fun2 %>% fun3`, R reads it as this: `((my_variable %>% fun1) %>% fun2) %>% fun3`.

So we repeatedly check that the current function/operator/call name is a pipe, if it is, grab the second entry (which is what is being piped into the current pipe). If it isn't, we've dug down far enough.

So, now that we have that little function, we can re-write our function to check for this first:
```{r get_name_def}
get_name <- function(x){
  lhs <- get_lhs()
  if(is.null(lhs)){
    lhs <- rlang::ensym(x)
  }
  as_name(lhs)
}
get_name(my_variable)
my_variable %>% get_name
```

Eureka! Now, let's check the wrapper function:
```{r wrap_get_name_lhs}
wrap_get_name <- function(wrapped_var){
  this_variable <- wrapped_var+1
  this_variable %>% get_name
}

wrap_get_name(my_variable)
my_variable %>% wrap_get_name
```

This function acts a little strange around `fseq` functions. But, the results make sense when you think about it.

```{r fseq_get_name_def}
fseq_get_name <- . %>% get_name
```

This method of creating a function, where the initial starting value of the pipeline is actually that previously discussed `.`, this is essentially the same as the previous, `wrap_get_name()` function:
```{r fseq_get_name_dummy, eval=F}
fseq_get_name_dummy <- function(.){
  . %>% get_name
}
```

As usual, we can use this function in one of two ways, either as a regular function with brackets or as a piped function

```{r fseq_get_name_brackets}
fseq_get_name(my_variable)
```
Looking at the alternate definition above, this makes sense as a result. The pipeline starts with a `.`

```{r fseq_get_name_pipe}
my_variable %>% fseq_get_name
```

What? This time, it's returned the value being piped in. But, if we imagine a `fseq`-style as sticking the pipelines together, then the actual start of this pipeline *is* the `my_variable`.


And there we go. You've now got a bit more of an understanding of the `sys.calls()` function and can extract the name of a variable being passed into a pipeline. This is a very basic way of doing it, it doesn't do nearly enough checks as a function in-production would have to do, but it's a good start. You could also extract any part of that original pipeline call. 


<p style="font-size: calc(100% - 6px);">Sidenote for the pro's out there. This page is written in `{rmarkdown}` and rendered using `{blogdown}`. This meant that when I used the `sys.calls()` function, I actually got a much deeper nesting of calls when rendering in these than in my RStudio application. This is because when rendering, each code chunk is evaluated *within another call*. For the local render with `{rmarkdown}`, I had to remove the first 18 calls before the "first" call was the one actually used above. For the `{blogdown}` render, it is 24. This page has a bunch of hidden code chunks (using `eval=F` and `echo=F`) to make the code and the output look seamless.</p>












