---
title: The lapply() family
author: Michael Barrowman
date: '2020-12-06'
slug: the-lapply-family
categories: []
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2020-12-06T15:05:08+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[This question on Reddit](https://www.reddit.com/r/rstats/comments/l0gvh3/good_site_to_understand_the_apply_collection_of/), got me thinking about the `lapply()` family of functions, and how a beginner might want to learn about them. Here is my take

# Introduction

 The easiest one to understand is `lapply()`, I'll work through that and then extend to the others. As an aside, the programmatic terminology is _vectorising_ as it allows us to perform an action over an entire _vector_ at once or _list_ in R.

Ignoring the dots, `lapply()` takes two arguments `X` and `FUN`. `FUN` is the name of the function, and `X` is a list of objects. When I say list, this could be an actual, as created by the `list()` function, or a vector such as `1:10`. But if you try to put something more complicated in like a `data.frame()`, you can get unexpected results (I'll come back to this).

# lists

So, let's say we have 

```{r}
X <- list(1:10,11:20,21:30)
X
```

This list has three elements, and each element consists of a vector of 10 numbers. We can access them using `[[`, where `X[[1]]` will return the first element, the numbers 1 to 10:
```{r}
X[[1]]
```

`X[[2]]` will return the second element, etc... This is _extraction_ as it extracts an element from a list. Extraction can only bring out a single element. We can also _subset_ using `[` for example 

```{r}
X[1:2]
```

This return the first and second elements. `X[1]` will return a subset consisting of the first element.
```{r}
X[1]
```


What's the difference between `X[1]` and `X[[1]]`? Well, `X[1]` returns a list, which is just 1 element long, that element being a vector of the numbers from 1 to 10. `X[[1]]` returns the actual element at position 1.

```{r}
length(X[1])
length(X[[1]])
class(X[1])
class(X[[1]])
```

So `X[1]` is a list, just like `X` but is shorter, a subset, just like how `X[1:2]` is a subset with length 2. Whereas `X[[1]]` _is_ the first element of `X`. This is clearer if we try to add something to these two objects:
```{r, error=T}
X[[1]] + 3
X[1] + 3
```
Again, to stress the point. `X[1]` is _not_ a number, it is a list containing a single element. Since `X[1]` is a list, we can therefore _extract_ that first element from it:
```{r}
X[1][[1]]
```

The difference between a list and a vector is that a vector has to all be of the same type (e.g. all characters as in `c("a","b","c")` or all numbers as in `c(1,2,3)`, the `c()` function will _coerce_ them otherwise, so `c(1,"2",3)` will coerce to characters. But a list can all be different, so `list("hello",2,1:10)` has three elements. In fact lists can contain lists (nested lists) 
```{r}
Y <- list("hello",1:10,list("one","two","three"))
Y
```
has three elements. If you _extract_ the third element, 
```{r}
Y[[3]]
```
you get another list. If you _subset_ the third element, 
```{r}
Y[3]
```
you get a list with 1 element.

As far as nomenclature is concerned, a _vector_ is a type of _list_ which has the requirement that all entries be of the same type. You can even use extraction on a vector, 
```{r}
x <- 1:10
x[3]
x[[3]]
```
Although when dealing with a vector, the second version is much less common. The reason this works is that _extraction_ and _subsetting_ are essentially the same thing in a vector (because it will always return a vector, it just might be of length 1).

# lapply()
So, now that we know what a list is, we can look at what `lapply()` does to that list. If we supply a function, `lapply()` will run that function on every element in that list. The simplest example would be, using `X` from above, `lapply(X,mean)` will return a list with the `mean()` of every element in `X`. 

```{r}
lapply(X,mean)
```

Remember that the elements in `X` are the vectors of numbers, `1:10`, `11:20` and `21:30`. We've _applied_ the function to the _list_ a _list-apply_.

The function doesn't have to be one that is named, and we can supply a function in-line

```{r}
lapply(X, function(x) mean(x-5.5))
```
This applies the function `function(x) mean(x-5.5)` to every element in `X`. You could define this function outside of the `lapply()` function earlier, but there is no need if this is the only place we plan on using it. 

For future, when the `R 4.1` version is released, I believe this will be even easier with the shorthand `\()` syntax.

```
lapply(X, \(x) mean(x - 5.5))
```

So running `lapply(X,FUN)` is the same as running the following `for()` loop

```
output <- vector("list",length(X))
for(i in 1:length(X)){
  output[[i]] <- FUN(X[[i]])
}
```

Compare the previous code to this:
```{r}
output <- vector("list",length(X))
for(i in 1:length(X)){
  output[[i]] <- mean(X[[i]])
}
output
```

Notice that I've defined the `output <- vector("list",length(X))` before running the `for()` loop. This line basically makes an empty list of the defined length. This will come up when we move on from `lapply()`

# dots
One part of `lapply()` that I've ignored is the `...` dots argument. These are basically other arguments that you want passed on to your function. Whatever is in the dots, will be passed to every call to `FUN`, whether named or not:

```{r}
lapply(c("a","b","c"),paste,"2")
```

```{r}
lapply(list( 1:10, c(1,2,NA,4), 21:30), mean, na.rm=T)
```

Essentially, this runs the following loop:

```{r}
X <- list( 1:10, c(1,2,NA,4), 21:30)
output <- vector("list",3)
for(i in 1:3){
  output[[i]] <- mean(X[[i]],na.rm=T)
}
output
```

Hopefully that will be enough to understand `lapply()`. One unusual case is using `lapply()` on a `data.frame`-like structure. Now, a `data.frame` looks like a table, but it's actually a list, but the list is counter intuitive. Each element of the list is a _column_ in the `data.frame`. So, if you run the following, you would get a result that is only 4 elements long

```{r}
iris0 <- iris[,1:4]
lapply(iris0,mean)
```

You might think that this would work _across_ the rows of the `data.frame`, but it works _down_ the columns. Also note that these outputs are now also named the same as the input list. This can be useful for keeping track of your inputs and outputs.

# apply()
This brings us only `apply()`.

The `apply()` function does a similar job, however it doesn't work on lists, it works on multi-dimensional objects, so matrices and arrays. It tries to collapse a multi-dimensional object down by one (or more) of its dimensions. So it turns a matrix into a vector (or an array into a smaller array). As well as `X` (which must be multi-dimensional, so definitely _not_ a list) and `FUN`, it also takes `MARGIN` which tells `apply()` which dimension(s) to collapse:

```{r}
M <- matrix(1:9,nrow=3)
apply(M,1,mean) #takes the mean of each row
apply(M,2,mean) #takes the mean of each column
```

The type returned is the same as the type we started with, and once again `apply()` can take other arguments as dots. So this works quite well with character matrices:
{
```{r}
M <- matrix(letters[1:9],nrow=3)
apply(M,1,paste0,collapse="") #pastes across the rows
apply(M,2,paste0,collapse="") #pastes down the columns
```

This means we can use `apply()` on a `data.frame` to work across the rows, rather than down the columns. In this case, ever though a `data.frame` is a list, because it can be accessed in the same way as a matrix, it still works

```{r}
apply(iris0,1,mean)
```

now gives a vector of the averages of each row.

# Rest of the family
Now for `lapply()`'s sisters:

`vapply()` takes an extra argument, which is of the same type as what you want your outcome to be. This is the one that I use most often. You can think of it like a `lapply()` that will output something other than a list. I usually give `FUN.VALUE` as something like `integer(1)` or `character(1)`. These functions generate empty vectors of that type, they are wrappers around things like `vector("integer",1)`

```{r}
X <- list(1:10,11:20,21:30)
vapply(X,mean,numeric(1))
```

This time, we get a numeric vector, rather than a list like we would with `lapply()`. I find this much easier to ensure I'm working with the correct type of data.

`sapply()` tries to simplify your output, So if `lapply()` outputs a list of vectors that are all the same length, instead of a list, it'll return a matrix

```{r}
X <- list(1:5, 6:10, 11:15)
sapply(X,range)
```

Each column in this result is the same as one of the elements of the list `lapply(X,range)`. They've just been `cbind`'d together. The use of `sapply()` is _not_ common as the output can be inconsistent. ``vapply()` is much prefered as it gives more control over the output. The above can be replicated with `vapply()` and will throw an error if the output is unexpected:

```{r}
X <- list(1:5, 6:10, 11:15)
vapply(X,range,numeric(2))
```
`tapply()` is more complicated as it subsets the `X` based on the `INDEX`. It describes this as a _"Ragged Array"_. I have _never_ used this directly, as I will usually do the subsetting manually using `split()`, but that is essentialy what `tapply()` does behind the scenes. `tapply()` also comes with a `simplify` argument, which decides whether R will try and simplify the results, like in `sapply()` or not, by default it will try and invoke this simplification. The following are therefore (roughly) equivalent

```
tapply(X, INDEX, FUN, simplify=FALSE)

lapply(split(X,INDEX), FUN)
```

`split()` creates a list where the first vector is split into groups based on the second argument.

So we can compare using both a `lapply()` and a `vapply()` 
```{r}
x <- 1:10
grp <- c(1,1,1,2,2,3,3,3,4,5)
tapply(x,grp,sum,simplify=FALSE)
lapply(split(x,grp),sum)

tapply(x,grp,sum)
vapply(split(x,grp),sum,numeric(1))

```

The other member of the `lapply()` family is `mapply()`. This is even more powerful as it allows you to _vectorise_ over multiple arguments, rather than just the first. Syntactically, the difference here is that the dots are the vectorised arguments, and the non-vectorised arguments go into the `MoreArgs` argument.

```{r}
X <- list("one","two",c("three", "four"))
Y <- list("A","B",c("C","D"))
mapply(paste,X,Y)
```

This is the same as doing:

```{r}
list(
  paste(X[[1]],Y[[1]]),
  paste(X[[2]],Y[[2]]),
  paste(X[[3]],Y[[3]])
)
```
Here is one final example using `rep()`, which repeats the first argument a specific number of times

```{r}
  X <- letters[1:4]
  Y <- 1:4
  mapply(rep,X,Y)
```
